# Modeling {#model}

## Libraries 

```{r}
# Libraries to install
# List of packages to check/install
packages <- c("data.table", "tidyverse", "rjags", "MCMCvis")

# Function to check and install packages
install_if_missing = function(pkg) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  } else {
    cat(paste("Package", pkg, "is already installed.\n"))
  }
}

# Apply the function to each package in the list
lapply(packages, install_if_missing)


# devtools::install_github("irap93/Bayesdemo") # Uncomment out this line install for data 

# Libraries
library(Bayesdemo)
library(data.table)
library(tidyverse)
library(rjags)
library(MCMCvis)
```
## Data
We have a dataset representing heifer performance, feeding behavior, and dry matter intake over a 56 day back grounding period. Heifers were assigned to a 2x2 crossover design to test a feed additives effect on animal behavior, performance, and morbidity during weaning.

```{r heifer}
data(heifer)
head(heifer)
summary(heifer)
```

## Excercise 1  



```{r heiferdata}
names(heifer)
ggplot(heifer, aes(x=cvDMI, y=Day56_ADG))+
  geom_point()+
  geom_smooth(method = 'lm')
```
Next we need to define the Jags model
```{r jagsmodel}
# Data list ----
# Data
data_list <- list(
  x = heifer$cvDMI,  # Example predictor values
  y = heifer$Day56_ADG, # Example response values
  N = nrow(heifer)     # Number of observations
)
## Jags model ----
# Define the model
model_string <- "
model {
  for (i in 1:N) {
    y[i] ~ dnorm(mu[i], tau)
    mu[i] <- beta0 + beta1 * x[i]
  }
  
  # Priors
  beta0 ~ dunif(-10, 10)
  beta1 ~ dunif(-1, 1)
  sigma ~ dunif(0, 100)
  tau <- 1 / (sigma * sigma)
}
"

# Initial values
inits <- list(
  list(beta0 = 0, beta1 = 0, sigma = 1)
)

# Parameters to monitor
params <- c("beta0", "beta1", "sigma", 'y')
```

Now we have the model set up, we are ready to run the model through jags
```{r runjags}
# Run the model
model <- jags.model(textConnection(model_string), data = data_list, inits = inits, n.chains = 1)
update(model, 1000)  # Burn-in
samples <- coda.samples(model, variable.names = params, n.iter = 5000)
```

And do a preliminary look at the results.

```{r}
beta0chain = MCMCchains(samples, params = 'beta0')
plot(density(beta0chain), main = "beta0chain")
rug(beta0chain, co = 'red')
abline(v = quantile(beta0chain, c(.025, .975)), col = "red", lwd = 2)
quantile(beta0chain, c(.025, .975))
mean(beta0chain)

beta1chain = MCMCchains(samples, params = 'beta1')
plot(density(beta1chain), main = 'beta1chain')
rug(beta1chain, col = 'blue')
abline(v = quantile(beta1chain, c(.025, .975)), col = "red", lwd = 2)
quantile(beta1chain, c(.025, .975))

y = MCMCpstr(samples, params = 'y')
plot(density(y$y), main = 'y')
rug(y$y, col = 'blue')
abline(v = quantile(y$y, c(.025, .975)), col = "red", lwd = 2)
quantile(y$y, c(.025, .975))
```
Summarise results
```{r}
# Summarize the results
summary(samples)
```

### Summary Plots
Prior Plots
```{r}
## Plots
# Prior plots ------
iter = 5000
priors = data.table(b0 = runif(iter,-10, 10),
                    b1 = runif(iter,-1, 1),
                    sigma = runif(iter, 0,100))
priors
class(samples)

summary(lm(data_list$y ~ data_list$x))
```

Posterior Plots
```{r}
## Beta 0 ----
observed = rnorm(iter, 1.01, 0.20)
beta0 = MCMCchains(samples, params = c("beta0"))
length(beta0)

plot(density(priors$b0), xlim = c(-.03, 3), ylim = c(0,3), col = 'green', main = 'Beta 0')
lines(density(beta0), col = 'blue')
lines(density(observed), col = 'red')

legend("topright", legend = c("Posterior", "Prior", "Scaled Likelihood"), 
       col = c("blue", "red", "green"), lwd = 2, lty = c(1, 2, 3))

# Beta 1 ----
uCV = mean(heifer$cvDMI)
sdCV = sd(heifer$cvDMI)
oCV = rnorm(iter, -0.211, 0.53)

beta1 = MCMCchains(samples, params = c("beta1"))
plot(density(priors$b1), xlim = c(-2, 3), ylim = c(0,1.5), col = 'green', main = 'Beta 1')
lines(density(beta1), col = 'blue')
lines(density(oCV), col = 'red')

legend("topright", legend = c("Posterior", "Prior", "Scaled Likelihood"), 
       col = c("blue", "red", "green"), lwd = 2, lty = c(1, 2, 3))
```

